---
title: "Association matrix as input"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{asso_as_input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(NetCoMi)
```

This demonstrates how to use the NetCoMi workflow when you already have an 
association matrix on which to base the network.

The QMP data set provided by the `SPRING` package is used to demonstrate how 
NetCoMi is used to analyze a precomputed network (given as association matrix).

The data set contains quantitative count data (true absolute values), which 
SPRING can deal with. See `?QMP` for details.

`nlambda` and `rep.num` are set to 10 for a decreased execution time, but should 
be higher for real data.

```{r association_input_1, message=FALSE}
library(SPRING)

# Load the QMP data set
data("QMP") 

# Run SPRING for association estimation
fit_spring <- SPRING(QMP, 
                     quantitative = TRUE, 
                     lambdaseq = "data-specific",
                     nlambda = 10, 
                     rep.num = 10,
                     seed = 123456, 
                     ncores = 1,
                     Rmethod = "approx",
                     verbose = FALSE)

# Optimal lambda
opt.K <- fit_spring$output$stars$opt.index
    
# Association matrix
assoMat <- as.matrix(SpiecEasi::symBeta(fit_spring$output$est$beta[[opt.K]],
                                        mode = "ave"))
rownames(assoMat) <- colnames(assoMat) <- colnames(QMP)
```

The association matrix is now passed to `netConstruct` to start the usual 
NetCoMi workflow. Note that the `dataType` argument must be set appropriately.

```{r association_input_2, fig.height=6, fig.width=6}
# Network construction and analysis
net_asso <- netConstruct(data = assoMat,
                         dataType = "condDependence",
                         sparsMethod = "none",
                         verbose = 0)

props_asso <- netAnalyze(net_asso, clustMethod = "hierarchical")
```


```{r association_input_3, fig.height=18, fig.width=20}
plot(props_asso,
     layout = "spring",
     repulsion = 1.2,
     shortenLabels = "none",
     labelScale = TRUE,
     rmSingles = TRUE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 2,
     nodeColor = "cluster",
     hubBorderCol = "gray60",
     cexNodes = 1.8,
     cexLabels = 2,
     cexHubLabels = 2.2,
     title1 = "Network for QMP data", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```
