---
title: "Cross-domain networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cross_domain_network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(NetCoMi)
```


This tutorial shows how to construct a cross-domain network (e.g. a network consisting of bacteria and fungi) using SpiecEasi's ability to estimate cross-domain associations.

We use the same data as in the "Cross domain interactions" section of the [SpiecEasi tutorial](https://github.com/zdk123/SpiecEasi).

The samples are split into two groups and cross-domain associations are computed for each group using SpiecEasi. The association matrices are then passed to NetCoMi's netConstruct() function to conduct a network comparison between the two groups.

**Note:**  
This tutorial explains how two cross-domain networks are constructed and **compared**. For constructing a single network, skip the step where the data are split into two groups and perform the framework only for a single data set (i.e. pass the estimated association matrix to the "data" argument of netConstruct() and continue with NetCoMi's standard pipeline).


```{r}
library(SpiecEasi)
library(phyloseq)

data(hmp2)

# Store count matrices (taxa are columns)
counts_hmp216S <- as.matrix(t(phyloseq::otu_table(hmp216S)@.Data))
counts_hmp2prot <- as.matrix(t(phyloseq::otu_table(hmp2prot)@.Data))

# Assume, the first 23 samples are in one group and the remaining 24 samples in the other group
group_vec <- c(rep(1, 23), rep(2, 24))

# Split count matrices
counts_hmp216S_gr1 <- counts_hmp216S[group_vec == 1, ]
counts_hmp216S_gr2 <- counts_hmp216S[group_vec == 2, ]

counts_hmp2prot_gr1 <- counts_hmp2prot[group_vec == 1, ]
counts_hmp2prot_gr2 <- counts_hmp2prot[group_vec == 2, ]

set.seed(123456)

# Run SpiecEasi and create association matrix for group 1
# Note: Increase nlambda and rep.num for real data sets
spiec_result_gr1 <- multi.spiec.easi(list(counts_hmp216S_gr1, 
                                          counts_hmp2prot_gr1), 
                                     method='mb', 
                                     nlambda=10, 
                                     lambda.min.ratio=1e-2, 
                                     pulsar.params = list(thresh = 0.05,
                                                          rep.num = 10))

assoMat1 <- SpiecEasi::symBeta(SpiecEasi::getOptBeta(spiec_result_gr1), mode = "ave")

assoMat1 <- as.matrix(assoMat1)

# Run SpiecEasi and create association matrix for group 2
# Note: Increase nlambda and rep.num for real data sets
spiec_result_gr2 <- multi.spiec.easi(list(counts_hmp216S_gr2, 
                                          counts_hmp2prot_gr2), 
                                     method='mb', 
                                     nlambda=10,
                                     lambda.min.ratio=1e-2, 
                                     pulsar.params = list(thresh = 0.05,
                                                          rep.num = 10))


assoMat2 <- SpiecEasi::symBeta(SpiecEasi::getOptBeta(spiec_result_gr2), mode = "ave")

assoMat2 <- as.matrix(assoMat2)

# Get taxa names
taxnames <- c(taxa_names(hmp216S), taxa_names(hmp2prot))

colnames(assoMat1) <- rownames(assoMat1) <- taxnames
diag(assoMat1) <- 1

colnames(assoMat2) <- rownames(assoMat2) <- taxnames
diag(assoMat2) <- 1

```

```{r}
# NetCoMi workflow 

library(NetCoMi)

# Network construction (pass association matrices to netConstruct)
# - sparsMethod must be set to "none" because sparsification is already included in SpiecEasi
net_hmp_16S_prot <- netConstruct(data = assoMat1, data2 = assoMat2, 
                                 dataType = "condDependence", 
                                 sparsMethod = "none")

# Network analysis
netprops_hmp_16S_prot <- netAnalyze(net_hmp_16S_prot, hubPar = "eigenvector")

```



```{r network_plot, fig.height=19, fig.width=40}

nodeCols <- c(rep("lightblue", ntaxa(hmp216S)), rep("orange", ntaxa(hmp2prot)))
names(nodeCols) <- taxnames

plot(netprops_hmp_16S_prot, 
     sameLayout = TRUE, 
     layoutGroup = "union",
     nodeColor = "colorVec", 
     colorVec = nodeCols,
     nodeSize = "eigen", 
     nodeSizeSpread = 2,
     labelScale = FALSE,
     cexNodes = 2, 
     cexLabels = 2,
     cexHubLabels = 2.5,
     cexTitle = 3.8,
     groupNames = c("group1", "group2"))


legend(-0.2, 1.2, cex = 3, pt.cex = 4, 
       legend = c("HMP2 16S", "HMP2 protein"), col = c("lightblue", "orange"), 
       bty = "n", pch = 16) 

```

```{r}
# Network comparison
# - Permutation tests cannot be performed because the association matrices are
#   used for network construction. For permutation tests, however, the count 
#   data are needed.
netcomp_hmp_16S_prot <- netCompare(netprops_hmp_16S_prot, permTest = FALSE)

summary(netcomp_hmp_16S_prot, groupNames = c("group1", "group2"))
```

