---
title: "Differential networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diffnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(NetCoMi)
```

A further possibility to compare two networks are so called 
*differential networks*, where two nodes are connected
if they are differentially associated between the two groups.

We start by dividing the amgut data set into two groups: samples with and 
without seasonal allergies. 

```{r}
data("amgut2.filt.phy")

# Split the phyloseq object into two groups
amgut_season_yes <- phyloseq::subset_samples(amgut2.filt.phy, 
                                             SEASONAL_ALLERGIES == "yes")
amgut_season_no <- phyloseq::subset_samples(amgut2.filt.phy, 
                                            SEASONAL_ALLERGIES == "no")

amgut_season_yes
amgut_season_no
```

Due to its very short execution time, we use Pearson's correlations for 
estimating associations between OTUs.

Fisher's z-test is applied for identifying differentially correlated OTUs. 
Multiple testing adjustment is done by controlling the local false discovery rate.

Note: `sparsMethod` is set to `"none"`, just to be able to include all 
differential associations in the association network plot (see below). 
However, the differential network is always based on the estimated association 
matrices before sparsification (the `assoEst1` and `assoEst2` matrices returned 
by `netConstruct()`).

```{r diffnet_1, fig.height=15, fig.width=30}
net_season_pears <- netConstruct(data = amgut_season_no, 
                                 data2 = amgut_season_yes, 
                                 filtTax = "highestVar",
                                 filtTaxPar = list(highestVar = 50),
                                 measure = "pearson", 
                                 normMethod = "clr",
                                 sparsMethod = "none", 
                                 thresh = 0.2,
                                 verbose = 3)

# Differential network construction
diff_season <- diffnet(net_season_pears,
                       diffMethod = "fisherTest", 
                       adjust = "lfdr")

# Differential network plot
plot(diff_season, 
     cexNodes = 0.8, 
     cexLegend = 3,
     cexTitle = 4,
     mar = c(2,2,8,5),
     legendGroupnames = c("group 'no'", "group 'yes'"),
     legendPos = c(0.7,1.6))

```

In the differential network shown above, edge colors represent the direction of 
associations in the two groups. If, for instance, two OTUs are positively 
associated in group 1 and negatively associated in group 2 (such as '191541' 
and '188236'), the respective edge is colored in cyan.  

We also take a look at the corresponding associations by constructing 
association networks that include only the differentially associated OTUs.

```{r diffnet_2, fig.height=6, fig.width=6}
props_season_pears <- netAnalyze(net_season_pears, 
                                 clustMethod = "cluster_fast_greedy",
                                 weightDeg = TRUE,
                                 normDeg = FALSE,
                                 gcmHeat = FALSE)
```


```{r diffnet_3, fig.height=19, fig.width=40}
# Identify the differentially associated OTUs
diffmat_sums <- rowSums(diff_season$diffAdjustMat)
diff_asso_names <- names(diffmat_sums[diffmat_sums > 0])

plot(props_season_pears, 
     nodeFilter = "names",
     nodeFilterPar = diff_asso_names,
     nodeColor = "gray",
     highlightHubs = FALSE,
     sameLayout = TRUE, 
     layoutGroup = "union",
     rmSingles = FALSE, 
     nodeSize = "clr",
     edgeTranspHigh = 20,
     labelScale = FALSE,
     cexNodes = 1.5, 
     cexLabels = 3,
     cexTitle = 3.8,
     groupNames = c("No seasonal allergies", "Seasonal allergies"),
     hubBorderCol  = "gray40")

legend(-0.15,-0.7, title = "estimated correlation:", legend = c("+","-"), 
       col = c("#009900","red"), inset = 0.05, cex = 4, lty = 1, lwd = 4, 
       bty = "n", horiz = TRUE)
```

We can see that the correlation between the aforementioned OTUs '191541' and 
'188236' is strongly positive in the left group and negative in the right group.
