---
title: "Dissimilarity-based Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dissimilarity_networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(NetCoMi)
```


If a **dissimilarity** measure is used for network construction, **nodes are subjects** 
instead of OTUs. The estimated dissimilarities are transformed into similarities,
which are used as edge weights so that subjects with a similar microbial 
composition are placed close together in the network plot. 

We construct a single network using Aitchison's distance being suitable for the 
application on compositional data. 

Since the Aitchison distance is based on the clr-transformation, zeros in the 
data need to be replaced. 

The network is sparsified using the k-nearest neighbor (knn) algorithm.

```{r example13}
data("amgut1.filt")

net_diss <- netConstruct(amgut1.filt,
                         measure = "aitchison",
                         zeroMethod = "multRepl",
                         sparsMethod = "knn",
                         kNeighbor = 3,
                         verbose = 3)
```

For cluster detection, we use hierarchical clustering with average linkage. 
Internally, `k=3` is passed to [`cutree()`](https://www.rdocumentation.org/packages/dendextend/versions/1.13.4/topics/cutree) 
from `stats` package so that the tree is cut into 3 clusters.

```{r example14, fig.height=6, fig.width=6}
props_diss <- netAnalyze(net_diss,
                         clustMethod = "hierarchical",
                         clustPar = list(method = "average", k = 3),
                         hubPar = "eigenvector")
```


```{r example15, fig.height=18, fig.width=20}
plot(props_diss, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     hubTransp = 40,
     edgeTranspLow = 60,
     charToRm = "00000",
     shortenLabels = "simple",
     labelLength = 6,
     mar = c(1, 3, 3, 5))

# get green color with 50% transparency
green2 <- colToTransp("#009900", 40)

legend(0.4, 1.1,
       cex = 2.2,
       legend = c("high similarity (low Aitchison distance)",
                  "low similarity (high Aitchison distance)"), 
       lty = 1, 
       lwd = c(3, 1),
       col = c("darkgreen", green2),
       bty = "n")

```

In this dissimilarity-based network, hubs are interpreted as samples with a 
microbial composition similar to that of many other samples in the data set. 
