---
title: "Network comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{netcompare}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(NetCoMi)
```

One of NetCoMi's strengths is the ability to compare networks between two groups. 
The `netCompare()` function is used for this task.

### Network construction

The amgut data set is split by `"SEASONAL_ALLERGIES"` leading to two subsets of 
samples (with and without seasonal allergies). We ignore the "None" group.  

```{r netcomp_spring_0}
data("amgut2.filt.phy")

# Split the phyloseq object into two groups
amgut_season_yes <- phyloseq::subset_samples(amgut2.filt.phy, 
                                             SEASONAL_ALLERGIES == "yes")
amgut_season_no <- phyloseq::subset_samples(amgut2.filt.phy, 
                                            SEASONAL_ALLERGIES == "no")

amgut_season_yes
amgut_season_no
```

The 50 nodes with highest variance are selected for network construction to get 
smaller networks.

We filter the 121 samples (sample size of the smaller group) with highest 
frequency to make the sample sizes equal and thus ensure comparability.

```{r netcomp_spring_1}
n_yes <- phyloseq::nsamples(amgut_season_yes)

# Network construction
net_season <- netConstruct(data = amgut_season_no, 
                           data2 = amgut_season_yes,  
                           filtTax = "highestVar",
                           filtTaxPar = list(highestVar = 50),
                           filtSamp = "highestFreq",
                           filtSampPar = list(highestFreq = n_yes),
                           measure = "spring",
                           measurePar = list(nlambda = 10, 
                                             rep.num = 10,
                                             Rmethod = "approx"),
                           normMethod = "none", 
                           zeroMethod = "none",
                           sparsMethod = "none", 
                           dissFunc = "signed",
                           verbose = 2,
                           seed = 123456)
```

Alternatively, a group vector could be passed to `group`, according to which
the data set is split into two groups:

```{r netcomp_spring_2, eval=FALSE}
# Get count table
countMat <- phyloseq::otu_table(amgut2.filt.phy)

# netConstruct() expects samples in rows
countMat <- t(as(countMat, "matrix"))

group_vec <- phyloseq::get_variable(amgut2.filt.phy, "SEASONAL_ALLERGIES")

# Select the two groups of interest (level "none" is excluded)
sel <- which(group_vec %in% c("no", "yes"))
group_vec <- group_vec[sel]
countMat <- countMat[sel, ]

net_season <- netConstruct(countMat, 
                           group = group_vec, 
                           filtTax = "highestVar",
                           filtTaxPar = list(highestVar = 50),
                           filtSamp = "highestFreq",
                           filtSampPar = list(highestFreq = n_yes),
                           measure = "spring",
                           measurePar = list(nlambda=10, 
                                             rep.num=10,
                                             Rmethod = "approx"),
                           normMethod = "none", 
                           zeroMethod = "none",
                           sparsMethod = "none", 
                           dissFunc = "signed",
                           verbose = 3,
                           seed = 123456)

```

### Network analysis

The object returned by `netConstruct()` containing both networks is again 
passed to `netAnalyze()`. Network properties are computed for both networks
simultaneously. 

To demonstrate further functionalities of `netAnalyze()`, we play around with 
the available arguments, even if the chosen setting might not be optimal. 

- `centrLCC = FALSE`: Centralities are calculated for all nodes (not only for the
largest connected component).
- `avDissIgnoreInf = TRUE`: Nodes with an infinite dissimilarity are ignored 
when calculating the average dissimilarity.
- `sPathNorm = FALSE`: Shortest paths are not normalized by average dissimilarity.
- `hubPar = c("degree", "eigenvector")`: Hubs are nodes with highest
degree and eigenvector centrality at the same time.
- `lnormFit = TRUE` and `hubQuant = 0.9`: A log-normal distribution is fitted to
the centrality values to identify nodes with "highest" centrality values. 
Here, a node is identified as hub if for each of the three centrality measures, 
the node's centrality value is above the 90% quantile of the fitted log-normal 
distribution.
- The non-normalized centralities are used for all four measures.

**Note! The arguments must be set carefully, depending on the research 
questions. NetCoMi's default values are not generally preferable in all 
practical cases!**

```{r netcomp_spring_3, fig.height=10, fig.width=10}
props_season <- netAnalyze(net_season, 
                           centrLCC = FALSE,
                           avDissIgnoreInf = TRUE,
                           sPathNorm = FALSE,
                           clustMethod = "cluster_fast_greedy",
                           hubPar = c("degree", "eigenvector"),
                           hubQuant = 0.9,
                           lnormFit = TRUE,
                           normDeg = FALSE,
                           normBetw = FALSE,
                           normClose = FALSE,
                           normEigen = FALSE)

summary(props_season)
```


### Visual network comparison

First, the layout is computed separately in both groups (qgraph's "spring" 
layout in this case).

Node sizes are scaled according to the mclr-transformed data since `SPRING` uses
the mclr transformation as normalization method.

Node colors represent clusters. Note that by default, two clusters have the same 
color in both groups if they have at least two nodes in common 
(`sameColThresh = 2`). Set `sameClustCol` to `FALSE` to get different cluster colors.

```{r netcomp_spring_4, fig.height=19, fig.width=40}
plot(props_season, 
     sameLayout = FALSE, 
     nodeColor = "cluster",
     nodeSize = "mclr",
     labelScale = FALSE,
     cexNodes = 1.5, 
     cexLabels = 2.5,
     cexHubLabels = 3,
     cexTitle = 3.7,
     groupNames = c("No seasonal allergies", "Seasonal allergies"),
     hubBorderCol  = "gray40")

legend("bottom", title = "estimated association:", legend = c("+","-"), 
       col = c("#009900","red"), inset = 0.02, cex = 4, lty = 1, lwd = 4, 
       bty = "n", horiz = TRUE)
```


Using different layouts leads to a "nice-looking" network plot for each group, 
however, it is difficult to identify group differences at first glance.

Thus, we now use the same layout in both groups. In the following, the layout is
computed for group 1 (the left network) and taken over for group 2.

`rmSingles` is set to `"inboth"` because only nodes that are unconnected in both
groups can be removed if the same layout is used.


```{r netcomp_spring_5, fig.height=19, fig.width=40}
plot(props_season, 
     sameLayout = TRUE, 
     layoutGroup = 1,
     rmSingles = "inboth", 
     nodeSize = "mclr", 
     labelScale = FALSE,
     cexNodes = 1.5, 
     cexLabels = 2.5,
     cexHubLabels = 3,
     cexTitle = 3.8,
     groupNames = c("No seasonal allergies", "Seasonal allergies"),
     hubBorderCol  = "gray40")

legend("bottom", title = "estimated association:", legend = c("+","-"), 
       col = c("#009900","red"), inset = 0.02, cex = 4, lty = 1, lwd = 4, 
       bty = "n", horiz = TRUE)
```

In the above plot, we can see clear differences between the groups. The OTU "322235", 
for instance, is more strongly connected in the "Seasonal allergies" group than
in the group without seasonal allergies, which is why it is a hub on the right, 
but not on the left.

However, if the layout of one group is simply taken over to the other, one of 
the networks (here the "seasonal allergies" group) is usually not that 
nice-looking due to the long edges. 
Therefore, NetCoMi (>= 1.0.2) offers a further 
option (`layoutGroup = "union"`), where a union of the two layouts is used 
in both groups.
In doing so, the nodes are placed as optimal as possible equally for both networks.

*The idea and R code for this functionality were provided by 
[Christian L. Müller](https://github.com/muellsen?tab=followers) and 
[Alice Sommer](https://www.iq.harvard.edu/people/alice-sommer)*


```{r netcomp_spring_6, fig.height=19, fig.width=40}
plot(props_season, 
     sameLayout = TRUE, 
     repulsion = 0.95,
     layoutGroup = "union",
     rmSingles = "inboth", 
     nodeSize = "mclr", 
     labelScale = FALSE,
     cexNodes = 1.5, 
     cexLabels = 2.5,
     cexHubLabels = 3,
     cexTitle = 3.8,
     groupNames = c("No seasonal allergies", "Seasonal allergies"),
     hubBorderCol  = "gray40")

legend("bottom", title = "estimated association:", legend = c("+","-"), 
       col = c("#009900","red"), inset = 0.02, cex = 4, lty = 1, lwd = 4, 
       bty = "n", horiz = TRUE)
```

### Quantitative network comparison

Since runtime is considerably increased if permutation tests are 
performed, we set the `permTest` parameter to `FALSE`. See the 
`tutorial_createAssoPerm` file for a network comparison including permutation 
tests.

Since permutation tests are still conducted for the Adjusted Rand Index, 
a seed should be set for reproducibility.

```{r netcomp_spring_7}
comp_season <- netCompare(props_season, 
                          permTest = FALSE, 
                          verbose = FALSE,
                          seed = 123456)

summary(comp_season, 
        groupNames = c("No allergies", "Allergies"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5)
```


