---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NetCoMi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The American Gut data provided by the [`SpiecEasi`](https://github.com/zdk123/SpiecEasi) package are used for almost all NetCoMi tutorials. 

We begin by constructing a single network, which we analyze using quantitative and graphical methods. Later, we will compare the networks of two groups: Individuals with and without seasonal allergies.

NetCoMi's main functions are `netConstruct()` for network construction, `netAnalyze()` for network analysis, and `netCompare()` for network comparison. As should become clear from the following examples, these three functions must be executed in the aforementioned order. A further function is `diffnet()` for constructing a differential association network. `diffnet()` must be applied to the object returned by `netConstruct()`.

First of all, we load NetCoMi and the data from American Gut Project (provided 
by [`SpiecEasi`](https://github.com/zdk123/SpiecEasi), which is automatically 
loaded together with NetCoMi).

```{r load_data, message=FALSE, warning=FALSE,}
library(NetCoMi)
data("amgut1.filt")
data("amgut2.filt.phy")
```

### Network construction 

We use the [SPRING](https://github.com/GraceYoon/SPRING) package for estimating associations (conditional dependence) between OTUs. 

The data are filtered within `netConstruct()` as follows:

- Only samples with a total number of reads of at least 1000 are included 
(argument `filtSamp`). 
- Only the 50 taxa with highest frequency are included (argument `filtTax`).

`measure` defines the association or dissimilarity measure, which is `"spring"` 
in our case. Additional arguments are passed to `SPRING()` via `measurePar`. 
`nlambda` and `rep.num` are set to 10 for a decreased execution time, but should 
be higher for real data. `Rmethod` is set to “approx” to estimate the correlations using a hybrid multi-linear interpolation approach proposed by @yoon2020fast. This method considerably reduces the runtime while controlling the approximation error.

Normalization as well as zero handling is performed internally in `SPRING()`. 
Hence, we set `normMethod` and `zeroMethod` to `"none"`.

We furthermore set `sparsMethod` to `"none"` because `SPRING` returns a sparse
network where no additional sparsification step is necessary.

We use the "signed" method for transforming associations into dissimilarities 
(argument `dissFunc`). In doing so, strongly negatively associated taxa have a 
high dissimilarity and, in turn, a low similarity, which corresponds to edge 
weights in the network plot.

The `verbose` argument is set to 3 so that all messages generated by 
`netConstruct()` as well as messages of external functions are printed.

```{r single_spring}
net_spring <- netConstruct(amgut1.filt,
                           filtTax = "highestFreq",
                           filtTaxPar = list(highestFreq = 50),
                           filtSamp = "totalReads",
                           filtSampPar = list(totalReads = 1000),
                           measure = "spring",
                           measurePar = list(nlambda=10, 
                                             rep.num=10,
                                             Rmethod = "approx"),
                           normMethod = "none", 
                           zeroMethod = "none",
                           sparsMethod = "none", 
                           dissFunc = "signed",
                           verbose = 2,
                           seed = 123456)
```

### Network analysis

NetCoMi's `netAnalyze()` function is used for analyzing the constructed 
network(s).

Here, `centrLCC` is set to `TRUE` meaning that centralities are calculated only 
for nodes in the largest connected component (LCC). 

Clusters are identified using greedy modularity optimization 
(by `cluster_fast_greedy()` from [`igraph`](https://igraph.org/r/) package).

Hubs are nodes with an eigenvector centrality value above the empirical 
95% quantile of all eigenvector centralities in the network (argument `hubPar`).

`weightDeg` and `normDeg` are set to `FALSE` so that the degree of a node is
simply defined as number of nodes that are adjacent to the node.

By default, a heatmap of the Graphlet Correlation Matrix (GCM) is returned (with 
graphlet correlations in the upper triangle and significance codes resulting 
from Student's t-test in the lower triangle). 
See `?calcGCM` and `?testGCM` for details.

```{r single_spring_2, fig.height=6, fig.width=6}
props_spring <- netAnalyze(net_spring, 
                           centrLCC = TRUE,
                           clustMethod = "cluster_fast_greedy",
                           hubPar = "eigenvector",
                           weightDeg = FALSE, 
                           normDeg = FALSE)

#?summary.microNetProps
summary(props_spring, numbNodes = 5L)
```

#### Plotting the GCM heatmap manually

```{r single_spring_heat, fig.height=6, fig.width=6}
plotHeat(mat = props_spring$graphletLCC$gcm1,
         pmat = props_spring$graphletLCC$pAdjust1,
         type = "mixed",
         title = "GCM", 
         colorLim = c(-1, 1),
         mar = c(2, 0, 2, 0))

# Add rectangles highlighting the four types of orbits
graphics::rect(xleft   = c( 0.5,  1.5, 4.5,  7.5),
               ybottom = c(11.5,  7.5, 4.5,  0.5),
               xright  = c( 1.5,  4.5, 7.5, 11.5),
               ytop    = c(10.5, 10.5, 7.5,  4.5),
               lwd = 2, xpd = NA)

text(6, -0.2, xpd = NA, 
     "Significance codes:  ***: 0.001;  **: 0.01;  *: 0.05")
```


### Visualizing the network

We use the determined clusters as node colors and scale
the node sizes according to the node's eigenvector centrality. 

```{r helppage, eval=FALSE}
# help page
?plot.microNetProps
```

```{r single_spring_3, fig.height=18, fig.width=20}
p <- plot(props_spring, 
          nodeColor = "cluster", 
          nodeSize = "eigenvector",
          title1 = "Network on OTU level with SPRING associations", 
          showTitle = TRUE,
          cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

Note that edge weights are (non-negative) similarities, however, the edges 
belonging to negative estimated associations are colored in red by default 
(`negDiffCol = TRUE`).

By default, a different transparency value is added to edges with an absolute 
weight below and above the `cut` value (arguments `edgeTranspLow` and 
`edgeTranspHigh`). The determined `cut` value can be read out as
follows:

```{r single_spring_4}
p$q1$Arguments$cut
```


### Export to Gephi

Some users may be interested in how to export the network to Gephi. Here's an example:

```{r}
# For Gephi, we have to generate an edge list with IDs.
# The corresponding labels (and also further node features) are stored as node list.

# Create edge object from the edge list exported by netConstruct()
edges <- dplyr::select(net_spring$edgelist1, v1, v2)

# Add Source and Target variables (as IDs)
edges$Source <- as.numeric(factor(edges$v1))
edges$Target <- as.numeric(factor(edges$v2))
edges$Type <- "Undirected"
edges$Weight <- net_spring$edgelist1$adja

nodes <- unique(edges[,c('v1','Source')])
colnames(nodes) <- c("Label", "Id")

# Add category with clusters (can be used as node colors in Gephi)
nodes$Category <- props_spring$clustering$clust1[nodes$Label]

edges <- dplyr::select(edges, Source, Target, Type, Weight)

write.csv(nodes, file = "nodes.csv", row.names = FALSE)
write.csv(edges, file = "edges.csv", row.names = FALSE)
```

The exported .csv files can then be imported into Gephi.

----

### Network with Pearson correlation as association measure

Let's construct another network using Pearson's correlation coefficient
as association measure. The input is now a `phyloseq` object.

Since Pearson correlations may lead to compositional effects when applied to 
sequencing data, we use the clr transformation as normalization method. Zero 
treatment is necessary in this case.

A threshold of 0.3 is used as sparsification method, so that only OTUs with an
absolute correlation greater than or equal to 0.3 are connected.


```{r single_pears_1, fig.height=18, fig.width=20}
net_pears <- netConstruct(amgut2.filt.phy,  
                          measure = "pearson",
                          normMethod = "clr",
                          zeroMethod = "multRepl",
                          sparsMethod = "threshold",
                          thresh = 0.3,
                          verbose = 3)
```

Network analysis and plotting:

```{r single_pears_2, fig.height=6, fig.width=6}
props_pears <- netAnalyze(net_pears, 
                          clustMethod = "cluster_fast_greedy")
```


```{r single_pears_3, fig.height=18, fig.width=20}
plot(props_pears, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     title1 = "Network on OTU level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:", 
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

Let's improve the visualization by changing the following arguments:

- `repulsion = 0.8`: Place the nodes further apart.
- `rmSingles = TRUE`: Single nodes are removed.
- `labelScale = FALSE` and `cexLabels = 1.6`: All labels have equal size and are
enlarged to improve readability of small node's labels.
- `nodeSizeSpread = 3` (default is 4): Node sizes are more similar if the value
is decreased. This argument (in combination with `cexNodes`) is useful to enlarge
small nodes while keeping the size of big nodes.
- `hubBorderCol = "darkgray"`: Change border color for a better readability of 
the node labels.

```{r single_pears_4, fig.height=18, fig.width=20}
plot(props_pears, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.8,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLabels = 1.6,
     nodeSizeSpread = 3,
     cexNodes = 2,
     hubBorderCol = "darkgray",
     title1 = "Network on OTU level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"),
       bty = "n", horiz = TRUE)
```

#### Edge filtering

The network can be sparsified further using the arguments `edgeFilter` (edges 
are filtered before the layout is computed) and `edgeInvisFilter` (edges are 
removed after the layout is computed and thus just made "invisible").

```{r single_pears_5, fig.height=18, fig.width=20}
plot(props_pears,
     edgeInvisFilter = "threshold",
     edgeInvisPar = 0.4,
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.8,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLabels = 1.6,
     nodeSizeSpread = 3,
     cexNodes = 2,
     hubBorderCol = "darkgray",
     title1 = paste0("Network on OTU level with Pearson correlations",
                     "\n(edge filter: threshold = 0.4)"),
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"),
       bty = "n", horiz = TRUE)
```

----

### Using the "unsigned" transformation

In the above network, the "signed" transformation was used to transform 
the estimated associations into dissimilarities. This leads to a network where 
strongly positive correlated taxa have a high edge weight (1 if the correlation 
equals 1) and strongly negative correlated taxa have a low edge weight (0 if 
the correlation equals -1).

We now use the "unsigned" transformation so that the edge weight between 
strongly correlated taxa is high, no matter of the sign. Hence, a correlation of 
-1 and 1 would lead to an edge weight of 1. 

#### Network construction

We can pass the network object from before to `netConstruct()` to save runtime.

```{r single_pears_unsigned_2, fig.height=18, fig.width=20}
net_pears_unsigned <- netConstruct(data = net_pears$assoEst1,
                                   dataType = "correlation", 
                                   sparsMethod = "threshold",
                                   thresh = 0.3,
                                   dissFunc = "unsigned",
                                   verbose = 3)
```

#### Estimated correlations and adjacency values

The following histograms demonstrate how the estimated correlations are 
transformed into adjacencies (= sparsified similarities for weighted networks).

Sparsified estimated correlations:

```{r single_pears_hist_1, fig.height=4, fig.width=8}
hist(net_pears$assoMat1, 100, xlim = c(-1, 1), ylim = c(0, 400),
     xlab = "Estimated correlation", 
     main = "Estimated correlations after sparsification")
```

Adjacency values computed using the "signed" transformation (values different 
from 0 and 1 will be edges in the network):

```{r single_pears_hist_2, fig.height=4, fig.width=8}
hist(net_pears$adjaMat1, 100, ylim = c(0, 400),
     xlab = "Adjacency values", 
     main = "Adjacencies (with \"signed\" transformation)")
```

Adjacency values computed using the "unsigned" transformation:

```{r single_pears_hist_3, fig.height=4, fig.width=8}
hist(net_pears_unsigned$adjaMat1, 100, ylim = c(0, 400),
     xlab = "Adjacency values", 
     main = "Adjacencies (with \"unsigned\" transformation)")
```

#### Network analysis and plotting

```{r single_pears_unsigned_3, fig.height=6, fig.width=6}
props_pears_unsigned <- netAnalyze(net_pears_unsigned, 
                                   clustMethod = "cluster_fast_greedy",
                                   gcmHeat = FALSE)
```


```{r single_pears_unsigned_4, fig.height=18, fig.width=20}
plot(props_pears_unsigned, 
     nodeColor = "cluster", 
     nodeSize = "eigenvector",
     repulsion = 0.9,
     rmSingles = TRUE,
     labelScale = FALSE,
     cexLabels = 1.6,
     nodeSizeSpread = 3,
     cexNodes = 2,
     hubBorderCol = "darkgray",
     title1 = "Network with Pearson correlations and \"unsigned\" transformation", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"),
       bty = "n", horiz = TRUE)
```

While with the "signed" transformation, positive correlated taxa are likely to 
belong to the same cluster, with the "unsigned" transformation clusters contain 
strongly positive and negative correlated taxa. 

----

### Network on genus level

We now construct a further network, where OTUs are agglomerated to genera. 

```{r single_genus_1, fig.height=6, fig.width=6}
library(phyloseq)
data("amgut2.filt.phy")

# Agglomerate to genus level
amgut_genus <- tax_glom(amgut2.filt.phy, taxrank = "Rank6")

# Taxonomic table
taxtab <- as(tax_table(amgut_genus), "matrix")

# Rename taxonomic table and make Rank6 (genus) unique
amgut_genus_renamed <- renameTaxa(amgut_genus, 
                                  pat = "<name>", 
                                  substPat = "<name>_<subst_name>(<subst_R>)",
                                  numDupli = "Rank6")

# Network construction and analysis
net_genus <- netConstruct(amgut_genus_renamed,
                          taxRank = "Rank6",
                          measure = "pearson",
                          zeroMethod = "multRepl",
                          normMethod = "clr",
                          sparsMethod = "threshold",
                          thresh = 0.3,
                          verbose = 3)

props_genus <- netAnalyze(net_genus, clustMethod = "cluster_fast_greedy")
```

#### Network plots

Modifications:

- Fruchterman-Reingold layout algorithm from `igraph` package used (passed to 
`plot` as matrix)
- Shortened labels (using the "intelligent" method, which avoids duplicates)
- Fixed node sizes, where hubs are enlarged
- Node color is gray for all nodes (transparancy is lower for hub nodes by default)

```{r single_genus_2, fig.height=18, fig.width=20}
# Compute layout
graph3 <- igraph::graph_from_adjacency_matrix(net_genus$adjaMat1, 
                                              weighted = TRUE)
set.seed(123456)
lay_fr <- igraph::layout_with_fr(graph3)

# Row names of the layout matrix must match the node names
rownames(lay_fr) <- rownames(net_genus$adjaMat1)

plot(props_genus,
     layout = lay_fr,
     shortenLabels = "intelligent",
     labelLength = 10,
     labelPattern = c(5, "'", 3, "'", 3),
     nodeSize = "fix",
     nodeColor = "gray",
     cexNodes = 0.8,
     cexHubs = 1.1,
     cexLabels = 1.2,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

Since the above visualization is obviously not optimal, we make further 
adjustments:

- This time, the Fruchterman-Reingold layout algorithm is computed within the 
plot function and thus applied to the "reduced" network without singletons
- Labels are not scaled to node sizes
- Single nodes are removed
- Node sizes are scaled to the column sums of clr-transformed data
- Node colors represent the determined clusters
- Border color of hub nodes is changed from black to darkgray
- Label size of hubs is enlarged

```{r single_genus_3, fig.height=18, fig.width=20}
set.seed(123456)

plot(props_genus,
     layout = "layout_with_fr",
     shortenLabels = "intelligent",
     labelLength = 10,
     labelPattern = c(5, "'", 3, "'", 3),
     labelScale = FALSE,
     rmSingles = TRUE,
     nodeSize = "clr",
     nodeColor = "cluster",
     hubBorderCol = "darkgray",
     cexNodes = 2,
     cexLabels = 1.5,
     cexHubLabels = 2,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

Let's check whether the largest nodes are actually those with highest 
column sums in the matrix with normalized counts returned by `netConstruct()`.

```{r single_genus_4}
sort(colSums(net_genus$normCounts1), decreasing = TRUE)[1:10]
```

In order to further improve our plot, we use the following modifications:

- This time, we choose the "spring" layout as part of `qgraph()` (the function 
is generally used for network plotting in NetCoMi)
- A repulsion value below 1 places the nodes further apart
- Labels are not shortened anymore
- Nodes (bacteria on genus level) are colored according to the respective phylum
- Edges representing positive associations are colored in blue, negative ones in
orange (just to give an example for alternative edge coloring)
- Transparency is increased for edges with high weight to improve the readability 
of node labels


```{r single_genus_5, fig.height=18, fig.width=20}
# Get phyla names
taxtab <- as(tax_table(amgut_genus_renamed), "matrix")
phyla <- as.factor(gsub("p__", "", taxtab[, "Rank2"]))
names(phyla) <- taxtab[, "Rank6"]
#table(phyla)

# Define phylum colors
phylcol <- c("cyan", "blue3", "red", "lawngreen", "yellow", "deeppink")

plot(props_genus,
     layout = "spring",
     repulsion = 0.84,
     shortenLabels = "none",
     charToRm = "g__",
     labelScale = FALSE,
     rmSingles = TRUE,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     posCol = "darkturquoise", 
     negCol = "orange",
     edgeTranspLow = 0,
     edgeTranspHigh = 40,
     cexNodes = 2,
     cexLabels = 2,
     cexHubLabels = 2.5,
     title1 = "Network on genus level with Pearson correlations", 
     showTitle = TRUE,
     cexTitle = 2.3)

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 60)

legend(-1.2, 1.2, cex = 2, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16) 

legend(0.7, 1.1, cex = 2.2, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("darkturquoise","orange"), 
       bty = "n", horiz = TRUE)
```

